Spá: a web-based viewer for text mining in Evidence Based Medicine

### 2014-05-13 doc/paper.tex
Remove reference to KDD submission

Summarizing the evidence about medical interventions is an immense undertaking, in part because unstructured \ac{pdf} documents remain the main vehicle for disseminating scientific findings.
Clinicians and researchers must therefore manually extract and synthesise information from these \acp{pdf}.
We introduce Spá,\footnote{From the Old Norse word spá or spæ referring to prophesying (prophecy)}\footnote{Source code available under GPLv3 at \url{https://github.com/joelkuiper/spa} \cite{Kuiper2014}; demo available at \url{http://spa.clinici.co/}} a web-based viewer that enables automated annotation and summarisation of \acp{pdf} via \acl{ml}.
To illustrate its functionality, we use Spá to semi-automate the assessment of bias in clinical trials.
%Spá can visualize the output from hybrid models that simultaneously classify documents (e.g., identifying trials as low or high risk of various biases), and annotate sentences that support these classifications.
Spá has a modular architecture, therefore the tool may be widely useful in other domains with a \ac{pdf}-based literature, including law, physics, and biology.

@@

As an example we consider risk of bias assessment.
Here reviewers assess, e.g., whether study participants and personnel were properly blinded \cite{Higgins2011}.
Assessing risk of bias is a time-consuming task.
A single trial typically takes a domain expert ten minutes \cite{Hartling2011}, and a single review typically includes several dozen trials.
Making matters worse, due to low rates of reviewer agreement it is regarded as best practice to have each study assessed twice by independent reviewers who later come to a consensus \cite{Hartling2009}.

@@

\begin{figure}[htb]
\vspace{-1em}
\centering
\includegraphics[width=0.8\linewidth]{./images/screenshot2.png}
\vspace{-1em}
\caption{\label{fig:screenshot}Screenshot of a \ac{pdf} with highlighted risk of bias. Here the risk of bias is assessed to be low, for example, and one of the supporting sentences for this assessment describes the randomization procedure (highlighted in green).}
\vspace{-1.5em}
\end{figure}

While our application of interest is \ac{ebm}, we emphasize that the visualization tool can be used for any domain in which one wants to annotate \acp{pdf}, e.g. genome-wide association studies or jurisprudence.
Thus the contribution of this work is two-fold, as we present:
(1) a practical tool that incorporates machine learning to help researchers rapidly assess the risk of biases in published biomedical articles, and,
(2) a general open-source system for visualizing the predictions of trained models from full-text articles on the web.
%These contributions are described further in sections \ref{section:EBM-ML} and \ref{section:architecture}.

>>>>

Summarizing the evidence about medical interventions is an immense undertaking, in part because unstructured \ac{pdf} documents remain the main vehicle for disseminating scientific findings.
Clinicians and researchers must therefore manually extract and synthesise information from these \acp{pdf}.
We introduce Spá\footnote{From the Old Norse word spá or spæ referring to prophesying (prophecy)}\footnote{Source code available under GPLv3 at \url{https://github.com/joelkuiper/spa} \cite{Kuiper2014}; demo available at \url{http://spa.clinici.co/}}, a web-based viewer that enables automated annotation and summarisation of \acp{pdf} via \acl{ml}.
To illustrate its functionality, we use Spá to semi-automate the assessment of bias in clinical trials.
%Spá can visualize the output from hybrid models that simultaneously classify documents (e.g., identifying trials as low or high risk of various biases), and annotate sentences that support these classifications.
Spá has a modular architecture, therefore the tool may be widely useful in other domains with a \ac{pdf}-based literature, including law, physics, and biology.

@@

As an example we consider risk of bias assessment.
Here reviewers assess, e.g., whether study participants and personnel were properly blinded \cite{Higgins2011}.
Assessing risk of bias is a time-consuming task.
A single trial may typically take a domain expert ten minutes \cite{Hartling2011}, and a single review may typically include several dozen trials.
Making matters worse, due to low rates of reviewer agreement it is regarded as best practice to have each study assessed twice by independent reviewers who later come to a consensus \cite{Hartling2009}.

@@

\begin{figure}[htb]
\vspace{-2em}
\centering
\includegraphics[width=0.8\linewidth]{./images/screenshot2.png}
\vspace{-1em}
\caption{\label{fig:screenshot}Screenshot of a \ac{pdf} with highlighted risk of bias. Here the risk of bias is assessed to be low, for example, and one of the supporting sentences for this assessment describes the randomization procedure (highlighted in green).}
\vspace{-1.5em}
\end{figure}

While our application of interest is \ac{ebm}, we emphasize that the visualization tool can be used for any domain in which one wants to annotate \acp{pdf}, e.g. genome-wide association studies or jurisprudence.
Thus the contribution of this work is two-fold, as we present:
(1) a practical tool that incorporates machine learning to help researchers rapidly assess the risk of biases in published biomedical articles, and,
(2) a general open-source web tool for visualizing the predictions of trained models from full-text articles.
%These contributions are described further in sections \ref{section:EBM-ML} and \ref{section:architecture}.
